Interfacing logic programming and Coq

### 2020-10-28 papers/types2020/cultures.tex
massaging Matteo's text for dep_pbt in the pbt section

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Recent
developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of ELPI with Coq.

>>>>

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Recent
developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of Elpi with Coq.

### 2020-10-28 papers/types2020/intro.tex
massaging Matteo's text for dep_pbt in the pbt section

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made ELPI available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions:  First,
ELPI adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of ELPI comes equipped with an API for accessing the
Coq environment and a quotation and anti-quotation syntax for mixing
for Coq expressions with \lP program elements.

The logic programming interpreter that underlies ELPI provides several
convenient features for the kind of meta-programming tasks that can
arise within modern proof assistants.  For example, \lP provides a
declarative and direct treatment of abstract syntax that contains
bindings, including capture-avoiding substitution,
unification, and recursive programming.  ELPI spares the
programmer from dealing with low-level aspects of bindings
representation (e.g., De Bruijn indexes) while still having clean and
effective ways to manipulate binding structures.

Since relations (not functions) play a central role in \lP, ELPI is
capable to providing direct support for the many relations that arise
in proof assistants.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics specifications
\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
operational semantics \cite{plotkin81,milner89book}).  Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using ELPI
\cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
higher-order logic programming as a meta-programming language useful
for Coq.

In this paper, we present two applications of ELPI within Coq. With
these examples, we shall illustrate that ELPI is useful not only
because of its meta-programming features,
% (e.g., a builtin treatment of bindings in syntax)
but also because it soundly
implements a higher-order intuitionistic logic.  The latter aspect of
ELPI makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.

Before we can present these examples, we first highlight the rather
striking differences in notions of computing and reasoning that arise
on each side of the ELPI-Coq API.   We will also present a quick
summary of the key proof theory concepts that are used by our example
applications.

@@

% LocalWords:  Sacerdoti Coen Tassi ELPI Teyjus Felty LCF tacticals
% LocalWords:  intuitionistic

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with an API for accessing the
Coq environment and a quotation and anti-quotation syntax for mixing
for Coq expressions with \lP program elements.

The logic programming interpreter that underlies Elpi provides several
convenient features for the kind of meta-programming tasks that can
arise within modern proof assistants.  For example, \lP provides a
declarative and direct treatment of abstract syntax that contains
bindings, including capture-avoiding substitution,
unification, and recursive programming.  Elpi spares the
programmer from dealing with low-level aspects of bindings
representation (e.g., De Bruijn indexes) while still having clean and
effective ways to manipulate binding structures.

Since relations (not functions) play a central role in \lP, Elpi is
capable to providing direct support for the many relations that arise
in proof assistants.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics specifications
\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
operational semantics \cite{plotkin81,milner89book}).  Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using Elpi
\cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
higher-order logic programming as a meta-programming language useful
for Coq.

In this paper, we present two applications of Elpi within Coq. With
these examples, we shall illustrate that Elpi is useful not only
because of its meta-programming features,
% (e.g., a builtin treatment of bindings in syntax)
but also because it soundly
implements a higher-order intuitionistic logic.  The latter aspect of
Elpi makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.

Before we can present these examples, we first highlight the rather
striking differences in notions of computing and reasoning that arise
on each side of the Elpi-Coq API.   We will also present a quick
summary of the key proof theory concepts that are used by our example
applications.

@@

% LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
% LocalWords:  intuitionistic

### 2020-10-28 papers/types2020/pbt.tex
massaging Matteo's text for dep_pbt in the pbt section

Let us consider a  relational specification of ordered list, as well as insertion in such lists:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, x > y -> insert x ys rs -> insert x (y :: ys) (x :: rs).

Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness. Since we did slip in an error, our tactic reports a counter-example, e.g.,
\verb|(* Proof Term: [0, [0; 1; 0]] *)|:
\begin{lstlisting}
Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs  Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded by
at most $5$ unfoldings of the library \lsti{Inductive} definition of
\lsti{nat} and \lsti{list}.

>>>>

\subsection{Implementing the tactic}
We will invoke the tactic in a
proof environment  where the goal is % one is trying to prove is the
property that the specification should meet. This means that, after
\lsti|intro| has been used to introduce the relevant hypotheses, we
will have a context % consisting of typing information for a bunch of variables
that we  divide in two categories: dependent
variables, which appear also in the types of other variables, and
non-dependent variables tat do not % Obviously, the
% two are the result of and \lsti|intro| on either a dependent or a
% non-dependent product
; in our setting this division acquires an additional reading beyond
what the property looks like: since specifications are encoded as pure
Horn clauses where no dependency is involved non-dependent variables
simply tag the specification clauses, whereas dependent variables
represent the typing information for the data variables that the
speficications use.

Whithin the tactic the user
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. The goal of the current
environment will directly provide the property for which we want to
find a counterexample. % Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}

To exemplify, let us add to the previous specification a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded
by at most $5$ unfoldings of the library \lsti{Inductive} definition
of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
tactic reports a counter-example, namely
\verb|(* Proof Term: [0, [0; 1; 0]] *)|

In order to generate the query to the \lsti|check| predicate, some
pre-processing is needed. In particular, we need to turn the data
variables into \lP logic variables when they appear inside a
specification, and to generate querys to \lsti|check| for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to code this preprocessing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses.
